Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-04 doc/paper-lcr.tex
more halo stuff

Define: convergence.

Where does it crop up? where is it important?

The convergence from an isolated mass sheet is the ratio of the projected surface mass density devided by the critical density at the redshift of the mass sheet.
\be
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distances from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is given by
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta})$ is the lens potential, derived from the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, or the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psikappa}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}).
\ee

@@

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW profiles \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
 and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are more concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. $\rho_0$ is calculated 
by the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee

From the NFW density profile, \citet{wrightbrainerd1999} derive the following quantities
which are necessary for our line of sight reconstruction:

The mass enclosed within radius is given by:

The projected mass density is given by:

The shear is given by:

>>>>

Convergence, or Ricci focusing, occurs when a gravitational lens focuses the rays in a raybundle.
This focussing can cause distant objects to appear brighter, and larger than they would if the lens
 were removed.
The convergence from an isolated mass sheet is the ratio of the projected surface mass density
(\Sigma) devided by the critical density at the redshift of the mass sheet (\Sigma_{\rm cr}),
\be
\kappa(\bmath{\theta})= \frac {{\Sigma} {\Sigma_{\rm cr}(z)}
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr}(z) \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee
and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$ are, respectively, the angular diameter distances from observer to the mass sheet, from observer to the source, and from the mass sheet to the source. If the surface mass density of the sheet exceeds $\Sigma_{\rm cr}$, multiple images of the source can
occur, otherwise only one image will be observed, although that image will still be perturbed
relative to the unlensed case.

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For an isolated lens, the time delay
function can be calculated from
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $\bmath{\theta}$ is the observed source position, $\bmath{\beta}$ is the 
unlensed source position, $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential. The Fermat potential is given by
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right], 
\ee
where $\psi(\bmath{\theta})$ is the lens potential, derived from the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, or the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psikappa}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}).
\ee

@@

With this catalogue of halos, positions, redshifts etc. it is possible to attempt
a reconstruction of the convergernce induced by every halo near a line of sight.
Cosmological dark matter simulations have shown that dark matter halos are well
approximated by NFW profiles \citep{NFW1997}, with a density profile given by
\be\label{eq:rhonfw}
\rho(r) = 
\frac{\rho_0}{\left(r/r_{s}\right)
\left(1+r/r_{s}\right)^{2}},
\ee
where $r_{s}$ is a characteristic radius of the cluster, representing where
the density slope transitions from $r^{-1}$ to $r^{-3}$. This radius is related
to the virial radius of the halo by $r_{s}~=~r_{200}/c$, where $r_{200}$ is the radius at 
which 
\be
\rho(r_{200}) = 200\rho_{{\mathrm{crit}}}(z), \text{ where } \rho_{\mathrm{crit}}(z) \equiv \frac{3H^{2}(z)}{8\pi G}
\ee
 and $c$ is the concentration parameter, which can be estimated from the halo's mass,
using a mass--concentration relation. Typically more massive halos are more concentrated,
but there is some scatter; we use the relation of \citet{neto2007} to estimate $c$
from the mass enclosed within $r_{200}$, which we denote as $M_{200}$. $\rho_0$ is calculated 
by the critical density and concentration parameter:
\be
\rho_0 = \rho_{\mathrm{crit}}(z)\frac{200}{3}\frac{c^{3}}{\ln(1+c)-c/(1+c)}.
\ee
Defining $x$ as the dimensionless projected radial distance $R/r_{s}$ , \citet{bartelmann1996} derives the projected mass density, which is given by:
\be\label{eq:kappanfw}
\Sigma_{\rm NFW}(x) = 
\begin{cases}
  \frac{2r_{s}\delta_{c}\rho_{c}}{x^{2}-1}
  \left(1-\frac{2}{\sqrt{1-x^{2}}}{\rm arctanh}\sqrt{\frac{1-x}{1+x}}\right)\hspace{0.15cm}
  & \left(x < 1\right) \\
\\
  \frac{2r_{s}\delta_{c}\rho_{c}}{3}
  & \left(x = 1\right)\\
\\
  \frac{2r_{s}\delta_{c}\rho_{c}}{x^{2}-1}
  \left(1-\frac{2}{\sqrt{x^{2}-1}}\arctan\sqrt{\frac{x-1}{1+x}}
  \right) 
  & \left(x > 1\right) 
\end{cases}
\ee
and the convergence is given by $\Sigma_{\rm NFW}(x) / \Sigma_{\rm cr}(z)$.

Distrubingly, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line sight will also be divergant\footnote{At large radius, $\Sigma_{\rm nfw} \propto R^{-2}$, but the differential number of halos centred within an annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$, so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which diverges logarithmically}.
Since infinite mass is unphysical, the profile must be truncated at some point. \flag{previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low \{}

%Hayashi2002 - truncation of satellites
